Development of restored dike grasslands to reference needs more than four years and can hardly modified by local site conditions

### 2023-02-10 manuscript.Rmd
Update manuscript.Rmd

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Seeding in general is a way to overcome the dispersal filter [@kiehl2006; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

@@

To measure the success of restoration, we chose four indices. First, we observed the persistence of the seeded species, which is a key interest of seeding as a restoration approach [@Wilsey.2021]. For this, it was surveyed if the seeded species are present in the respective plot in following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site from an ordination [@Rydgren.2019]. To complement the indices, the success of restoration or the difference from desired conditions needs to be evaluated [@Brudvig.2017b]. Therefore, the species composition of the experimental plots were compared with reference sites. We asked following question to evaluate the restoration approaches and success after four years:

@@

The seed mixture-substrate combination were tested with an experiment on a dike of the River Danube in SE Germany, which was established in March 2018 (Figure \@ref(fig:map)); 314 m above sea level; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix Table @ref(tab:flood-weather), @lfu2021waterlevel; @lfu2021discharge]. Regional sand (0--4 mm) was used to lean the substrate and the soil was taken from removed agricultural soil of a dike construction site near Steinkirchen. The substrates were mixed with a big roller and an excavator prepared the plots. The target vegetation were lowland hay meadows [EUNIS code: R22, @Chytry.2020; Arrhenatherion elatioris, EuroVegChecklist: CM01A, @mucina2016] and semi-dry calcareous grassland (R1A; Cirsio-Brachypodion pinnati, DA01B). The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From the species pools, 20 species were randomly selected for each plot. The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Figure \@ref(fig:traits)). As expected, the hay meadows mixtures have a higher specific leaf area (SLA) lower seed mass and higher canopy height than the dry meadows mixtures. The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by 'normal' deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was seeded as a nursery grass to provide safe sites under drought conditions.

@@

## The effect of seed mixture-substrate combinations

@@

sand admixture positive but 25% enough (ref. to chenot?)

@@

no significant effect of combination (variability important, further research)

@@

@@

```{r tab.id = "establishment", tab.cap = "Establishment rate of sown species [Nomenclature: @WorldFloraOnline.2021]. Indicated is the number of seeded plots and the percentage of these plots which were occupied by the respective species. This means that for each species only the seeded plots were included. The shades of grey indicate the success of establishment."}
read_csv(
  here("data", "processed", "data_processed_traits.csv"),
                 col_names = TRUE, na = c("na", "NA"), col_types =
                   cols(.default = "?", name = "f")
  ) %>%
  select(name, family, total_seeded_2018, starts_with("rate_")) %>%
  drop_na(total_seeded_2018) %>%
  arrange(match(family, c("Poaceae", "Fabaceae"))) %>%
  mutate(across(starts_with("rate"), ~ .x * 100),
         name = str_replace_all(name, "_", " ")) %>%
  rename(Species = name, Family = family, "Seeded plots [#]" = total_seeded_2018, "Establishment rate [%]" = rate_2018, "2019" = rate_2019, "2020" = rate_2020, "2021" = rate_2021) %>%
  flextable() %>%
  align(part = "all", align = "center", j = -1) %>%
  merge_at(part = "header", j = 4:7) %>%
  add_header_row(top = FALSE, values = c("", "", "", "2018", "2019", "2020", "2021")) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  bold(part = "header") %>%
  style(j = 1, pr_t = fp_text_default(italic = TRUE)) %>%
  bg(j = 4:7, bg = scales::col_numeric(palette = c("white", "grey60"), domain = c(0, 100))) %>%
  autofit()
```

>>>>

Sowing high-diversity seed mixtures of local provenance produced by commercial but specialized seed companies is a promising way to scale up restoration efforts. Though, there are still open questions about specifying the seed mixtures to specific site conditions or even more to future climate conditions to improve efficiency and effectiveness of restoration efforts [@Torok.2021]. For this reason, restoration ecology has to increase predictability of restoration approaches [@Suding.2011; @Mouquet.2015]. Rigorous, repeatable, and transparent experiments linked to a hypothesis derived from theory are needed to guide evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration approach are key predictors of the development of vegetation after sowing [@Brudvig.2017b]. Seeding in general is a way to overcome the dispersal filter [@myers2009; @orrock2023], but we focus on environmental and the biotic filtering which are the main assembly processes to be manipulated by the choice of the right seed mixture-substrate combination [@Torok.2017]. This means a mutually adaption of the substrate to the preferences of the target vegetation, and the seed mixtures to the characteristics of the chosen substrate. The right substrates should reduce habitat filtering and the right seed mixture should minimize competitive exclusion of unsuited species. Modifying seed mixtures for certain site conditions came up with functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020] but it is not easy to put it into practice for novel ecosystems like urban grasslands [@Bauer.2022]. Substrates cannot only be modified for conservation purposes in urban areas, but also in quarries [@chenot-lescure2022] or dikes [@Liebrand.1996].

@@

To measure the success of restoration, we chose four indices. First, we observed the persistence of the sown species, which is a key interest of seeding as a restoration approach [@Wilsey.2021]. For this, it was surveyed if the sown species are present in the respective plot in following years. Second, we calculated the Favorable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are target species for a habitat type of interest. Third, the successional distance to the reference was determined. This measure quantifies the distance of a plot to the average reference site from an ordination [@Rydgren.2019]. To complement the indices, the success of restoration or the difference from desired conditions needs to be evaluated [@Brudvig.2017b]. Therefore, the species composition of the experimental plots were compared with reference sites. We asked following question to evaluate the restoration approaches and success after four years:

@@

The seed mixture-substrate combination were tested with an experiment on a dike of the River Danube in SE Germany, which was established in March 2018 (Figure \@ref(fig:map)); 314 m above sea level; WGS84 (lat/lon), 48.83895, 12.88412). The climate is temperate-suboceanic with a mean annual temperature of 8.4 °C and an annual precipitation of 984 mm [@dwd2021]. During the study, three exceptionally dry years (2018--2020) occurred [@hari2020; @cdc2022temp; @cdc2022prec], and three minor floods which did not reach the plots [Appendix Table @ref(tab:flood-weather), @lfu2021waterlevel; @lfu2021discharge]. Regional sand (0--4 mm) was used to lean the substrate and the soil was taken from removed agricultural soil of a dike construction site near Steinkirchen. The substrates were mixed with a big roller and an excavator prepared the plots. The target vegetation were lowland hay meadows [EUNIS code: R22, @Chytry.2020; Arrhenatherion elatioris, EuroVegChecklist: CM01A, @mucina2016] and semi-dry calcareous grassland (R1A; Cirsio-Brachypodion pinnati, DA01B). The species pool of hay meadows and dry grasslands consisted of 55 and 58 species, respectively. The seeds were received from a commercial producer of autochthonous seeds. From the species pools, 20 species were randomly selected for each plot. The seed mixtures always contained seven grasses (60 wt% of total seed mixture), three legumes (5%) and ten herbs (35%) (Table \@ref(tab:seedmix)). The community-weighted means (CWM) of functional traits differed between the seed mixture types (Figure \@ref(fig:traits)). As expected, the hay meadows mixtures have a higher specific leaf area (SLA) lower seed mass and higher canopy height than the dry meadows mixtures. The south exposed plots were sown in mid-April 2018 and the north exposition two weeks later. Due to the drought, the south exposition was protected by a geotextile consisting of straw chaff (350 g m^-2^) which was removed after two weeks.

The management started with a cut without hay removal and a mowing height of 20 cm in August 2018, followed by 'normal' deep cuts with hay removal in July 2019 and 2020. In October 2018, *Bromus hordeaceus* was sown as a nursery grass to provide safe sites under drought conditions.

@@

## General effects of treatments and exposition

As expected, sand admixture led to higher persistence of sown species. In addition, 50% admixture did not further improve the result and the substrate depth had no effect. Restoration on agricultural soils has often limited success due to high nutrient loads [@walker2004]. Our results suggest a slightly better adaption of hay meadows to the north exposition than dry grasslands (Figure @ref(fig:recovery)). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north exposed dike slopes [@Batori.2020; @Oberdorfer.1993b].

The vegetation on the south exposition performed consistently and significantly worse than the north exposition. One reason might be the removal of the geotextile on the south exposition which led to the drying of many seedlings why the distinction of the seed mixture types was leveled. The removal was necessary because of the death of several protected lizards. Moreover, the intense drought of 2018 and 2019 [@hari2020] might have decreased germination and establishment of sown species because a moist spring with a mesic temperature improves establishment [@Larson.2021; @orrock2023]. The interesting outcome of this unfavorable situation during establishment on the south exposition is the lasting negative effect. This suggests a legacy effect by unfavorable weather conditions after sowing [@Stuble.2017; @Groves.2020] which might have led to an alternative trajectory and could probably only be met by resowing.

## No interaction effect of seed mixture-substrate combinations

@@

@@

no significant effect of combination (variability important, further research, chenot with topsoil)

@@

Plant traits (Plas 2020, Orrock 2023)

@@

```{r tab.id = "establishment", tab.cap = "Establishment rate of sown species [Nomenclature: @WorldFloraOnline.2021]. Indicated is the number of sown plots and the percentage of these plots which were occupied by the respective species. This means that for each species only the sown plots were included. The shades of grey indicate the success of establishment."}
read_csv(
  here("data", "processed", "data_processed_traits.csv"),
                 col_names = TRUE, na = c("na", "NA"), col_types =
                   cols(.default = "?", name = "f")
  ) %>%
  select(name, family, total_seeded_2018, starts_with("rate_")) %>%
  drop_na(total_seeded_2018) %>%
  arrange(match(family, c("Poaceae", "Fabaceae"))) %>%
  mutate(across(starts_with("rate"), ~ .x * 100),
         name = str_replace_all(name, "_", " ")) %>%
  rename(Species = name, Family = family, "Seeded plots [#]" = total_sown_2018, "Establishment rate [%]" = rate_2018, "2019" = rate_2019, "2020" = rate_2020, "2021" = rate_2021) %>%
  flextable() %>%
  align(part = "all", align = "center", j = -1) %>%
  merge_at(part = "header", j = 4:7) %>%
  add_header_row(top = FALSE, values = c("", "", "", "2018", "2019", "2020", "2021")) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  bold(part = "header") %>%
  style(j = 1, pr_t = fp_text_default(italic = TRUE)) %>%
  bg(j = 4:7, bg = scales::col_numeric(palette = c("white", "grey60"), domain = c(0, 100))) %>%
  autofit()
```
